Ball path curvature and in-game free throw shooting proficiency in the National Basketball Association

### 2025-06-06 reports/arxiv/paper.tex

\title{Ball path curvature and in-game free throw shooting proficiency in the National Basketball Association}
\author[1]{Judy Zhu}
\author[2]{Dave Love}
\author[3]{Scott Powers}
\affil[1]{Department of Mathematics, Rice University}
\affil[2]{CDL Basketball Enterprises}
\affil[3]{Department of Sport Management, Rice University}

@@

  \begin{abstract}
    Basketball shooting coaches agree that smoother shooting motions are better, but there is less agreement about what ``smooth'' means, quantitatively, or what part of the shooting motion needs to be smooth. Using ball tracking data from the 2023-2024 National Basketball Association regular season, we explore the relationship between ball path curvature and free throw shooting performance. We fit Bézier curves to the ball tracking data in the sagittal plane and test different methods of calculating path curvature. We find that both max curvature and terminal curvature are negatively associated with shooting performance, but terminal curvature explains much more of the between-player variance in free throw shooting performance. This suggests that shooting coaches would be better off focusing on the smoothness of the end of the shot rather than at the beginning of the forward motion of the ball.
  \end{abstract}

@@

    \citet{slegers_role_2024} proposed a framework for analyzing the free throw ball paths using curvature-based metrics and collected data from jump shots attempted by male professional basketball players in a controlled, experimental environment. Surprisingly, they found that max ball path curvature to be positively associated with shooting performance. Rather, terminal ball path curvature (more weight on curavture closer to release) was negatively associated with shooting performance. In other words, shooters should not want to smooth out the peak curvature---they should want to smooth out the curvature as the ball approaches release. The recommended focus on terminal path curvature makes intuitive sense because it means keeping the path as smooth as possible when the ball is moving fastest.

    The present study builds directly on the methodology of \citet{slegers_role_2024}, adapting to work with in-game data from the National Basketball Association (NBA) collected via Second Spectrum. This adaptation involves pre-processing noisy ball-tracking data, implementing ball path standardization methods for fair comparison across players, creating a  three-dimensional metric to address the limitations of curvature-based approaches that operate solely within the sagittal plane, and regressing the metrics directly against the players' season free throw shooting percentage (FT\%). In doing so, this study not only seeks to validate the findings of \citet{slegers_role_2024} in game conditions, but also contributes a robust and generalizable framework for evaluating shot smoothness and shooting accuracy at the elite level.

@@

      A similar concept of individualized optimization in shooting mechanics has been applied to release strategies, including release angles and velocity. Research has demonstrated that maximum shooting performance is achieved when athletes match their release parameters to their unique mechanics, rather than conforming to a predetermined ideal release angle \citep{slegers_basketball_2022}. These findings reinforce the idea that an optimal shot is not universal but highly specific to players, shaped by their experiences, physical attributes and muscle coordination.

@@

      In addition to release conditions, recent research has emphasized the role of ball trajectory and curvature in determining shooting accuracy. \citet{slegers_role_2024} revealed a negative correlation between the terminal curvature of the ball path and longitudinal shooting accuracy, with the latter being assessed using the standard deviation of intra-individual release velocity as a proxy, given its relevance to distance control. Furthermore, \citet{slegers_role_2024} validated that in the S-shaped shooting path, better shooters exhibit a sharper transition from the backswing to the forward shooting motion compared to less skilled players. Their study involved 31 professional male basketball players, each of whom attempted 40 jump shots under controlled conditions.

@@

    \begin{figure}[H]
        \centering
        \includegraphics[width=0.8\linewidth]{reports/arxiv/distributions.png}
        \caption{\it Distributions of key shot smoothness metrics and free throw percentage (FT\%) across all players in the study. KDE curves are used to illustrate the underlying shape of each distribution.}
        \label{fig:scattered}
    \end{figure}

@@

    Table \ref{tab:wls-results} summarizes the results of estimating the WLS model (\ref{eqn:weighted-least-squares}) for each of the four curvature metrics. Each metric is a statistically significant predictor of FT\%, with each $p$-value below 0.01. Among them, the time-weighted curvature integrals demonstrate the strongest explanatory power, particularly at $\ell = 5$, which achieves the highest $R^2$ value of 0.330. This finding aligns with the controlled experiments of \citet{slegers_role_2024}, suggesting that lower terminal curvature is correlated with greater shooting performance.

@@

>>>>

\title{Ball path curvature and in-game free throw shooting proficiency in the National Basketball Association}
\author[1]{Ruoqian Zhu}
\author[2]{Dave Love}
\author[3]{Scott Powers}
\affil[1]{Department of Mathematics, Rice University}
\affil[2]{CDL Basketball Enterprises}
\affil[3]{Department of Sport Management, Rice University}

@@

  \begin{abstract}
    Basketball shooting coaches agree that smoother shooting motions are better, but there is less agreement about what ``smooth'' means quantitatively or what part of the shooting motion needs to be smooth. Using ball tracking data from the 2023-2024 National Basketball Association regular season, we explore the relationship between ball path curvature and free throw shooting performance. We fit Bézier curves to the ball tracking data in the sagittal plane and test different methods of calculating path curvature. We find that both max curvature and terminal curvature are negatively associated with shooting performance, but terminal curvature explains much more of the between-player variance in free throw shooting performance. This suggests that shooting coaches would be better off focusing on the smoothness of the end of the shot rather than at the beginning of the forward motion of the ball.
  \end{abstract}

@@

    \citet{slegers_role_2024} proposed a framework for analyzing the free throw ball paths using curvature-based metrics and collected data from jump shots attempted by male professional basketball players in a controlled, experimental environment. Surprisingly, they found that max ball path curvature to be positively associated with shooting performance. Rather, terminal ball path curvature (more weight on curvature closer to release) was negatively associated with shooting performance. In other words, shooters should not want to smooth out the peak curvature---they should want to smooth out the curvature as the ball approaches release. The recommended focus on terminal path curvature makes intuitive sense because it means keeping the path as smooth as possible when the ball is moving the fastest.

    The present study builds directly on the methodology of \citet{slegers_role_2024}, adapting to work with in-game data from the National Basketball Association (NBA) collected via Second Spectrum. This adaptation involves pre-processing noisy ball-tracking data, implementing ball-path standardization methods for fair comparison across players, and regressing the metrics directly against the players' season free throw shooting percentage (FT\%). In doing so, this study not only seeks to validate the findings of \citet{slegers_role_2024} in game conditions, but also contributes a robust and generalizable framework for evaluating shot smoothness and shooting accuracy at the elite level.

@@

      A similar concept of individualized optimization in shooting mechanics has been applied to release strategies, including release angles and velocity. Research has shown that maximum shooting performance is achieved when athletes match their release parameters to their unique mechanics, rather than conforming to a predetermined ideal release angle \citep{slegers_basketball_2022}. These findings reinforce the idea that an optimal shot is not universal but highly specific to players, shaped by their experiences, physical attributes, and muscle coordination.

@@

      In addition to release conditions, recent research has emphasized the role of ball trajectory and curvature in determining shooting accuracy. \citet{slegers_role_2024} revealed a negative correlation between the terminal curvature of the ball path and longitudinal shooting accuracy, with the latter being assessed using the standard deviation of the intra-individual release velocity as a proxy, given its relevance to distance control. Furthermore, \citet{slegers_role_2024} validated that in the S-shaped shooting path, better shooters exhibit a sharper transition from the backswing to the forward shooting motion compared to less skilled players. Their study involved 31 professional male basketball players, each of whom attempted 40 jump shots under controlled conditions.

@@

    \begin{figure}[H]
        \centering
        \includegraphics[width=0.8\linewidth]{reports/arxiv/distributions.png}
        \caption{\it Distributions of key shot smoothness metrics across all players in the study. KDE curves are used to illustrate the underlying shape of each distribution.}
        \label{fig:scattered}
    \end{figure}

@@

    Table \ref{tab:wls-results} summarizes the results of estimating the WLS model (\ref{eqn:weighted-least-squares}) for each of the four curvature metrics. Each metric is a statistically significant predictor of FT\%, with each $p$-value below 0.01. Among them, the time-weighted curvature integrals demonstrate the strongest explanatory power, particularly at $\ell = 5$, which achieves the highest $R^2$ value of 0.330. The confidence intervals shown in Figure~\ref{fig:regressionlines} further supports this conclusion: the regression fits, particularly at $\ell = 3$ and $\ell = 5$, exhibit relatively narrow and stable bands, indicating greater precision in the estimated relationship between curvature and FT\% in these models. This finding aligns with the controlled experiments of \citet{slegers_role_2024}, suggesting that lower terminal curvature is correlated with greater shooting performance.

@@

    \begin{figure}[H]
        \centering
        \includegraphics[width=0.6\linewidth]{reports/arxiv/regressionlines.png}
        \caption{\it Weighted Least Squares (WLS) regression models illustrating the relationship between free throw percentage (FT\%) and four curvature-based shooting metrics. The black lines indicate the WLS regression fits, and the shaded red bands represent the 95\% confidence intervals around each regression line. Across all subplots, higher curvature values are associated with lower FT\%.}
        \label{fig:scattered}
    \end{figure}
